Mapping

### 2024-03-13 episodes/data-visualization.Rmd
Adaptation to Microbial Evolution and Epidemiology course; added pages, fused some

---
title: 'Data Visualization'
teaching: 10
exercises: 60
---

@@

- Explore the resulting trees in conjunction with the meta data.
- Make an estimation on the likelihood of transmission events

@@

- How are phylogenetic trees viewed and compared?
- How can I visualize several layers of data?

@@

@@

- Genetic information can confirm or contradict the meta data

>>>>

---
title: 'Molecular epidemiology'
teaching: 0
exercises: 240
---

The low mutation rate of *M. tuberculosis* means that even at the highest resolution provided by whole genome sequencing it is still difficult to confidently affirm the inferences of transmission made by traditional epidemiological techniques. This means it is very difficult to determine transmission inclusively. However, whole genome sequencing does in some cases allow us to exclude direct transmission, by using the phylogenetic context provided by other strains. Not only does whole genome sequencing provide the inter-cluster differentiation provided by current typing methods, but it also achieves intra-cluster resolution which can be used to inform epidemiological investigation.

@@

- Explore the resulting trees in conjunction with the meta data.
- Make an estimation on the likelihood of transmission events
- Compare and interpret the different data sources.

@@

- How are phylogenetic trees viewed and compared?
- How can I visualize several layers of data?
- In which cases is transmission likely?

@@

Compare the pangenome clustering with the clustering based on SNPs and with the data from
epidemiological contact tracing:

Cluster A

- Patient A1 - sample ERR029207
- Patient A2 - sample ERR029206
- Patient A3 - sample ERR026478

Cluster B

- Patient B1 - sample ERR026474
- Patient B2 - sample ERR026473

Cluster C

- Patient C1 - sample ERR026481
- Patient C2 - sample ERR026482

Which transmission events are affirmed by the genomic data? Which ones partially or not? Why?

@@

- SNP phylogeny and metadata can convey different messages
- Human interpretation is often needed to weigh the different information sources.
- The low mutation rate of *M. tuberculosis* does not allow to make confident inferences of transmission but does allow to exclude transmission

### 2024-03-13 episodes/msa.Rmd
Adaptation to Microbial Evolution and Epidemiology course; added pages, fused some

---
title: 'Multiple sequence alignment'
teaching: 10
exercises: 30
---

@@

::::::::::::::::::::::::::::::::::::: challenge

## Challenge 1: Can you do it?

What is the output of this command?

```r
paste("This", "new", "lesson", "looks", "good")
```

:::::::::::::::::::::::: solution

## Output

```output
[1] "This new lesson looks good"
```

:::::::::::::::::::::::::::::::::

## Challenge 2: how do you nest solutions within challenge blocks?

:::::::::::::::::::::::: solution

You can add a line with at least three colons and a `solution` tag.

:::::::::::::::::::::::::::::::::
::::::::::::::::::::::::::::::::::::::::::::::::

## Figures

You can include figures generated from R Markdown:

```{r pyramid, fig.alt = "pie chart illusion of a pyramid", fig.cap = "Sun arise each and every morning"}
pie(
  c(Sky = 78, "Sunny side of pyramid" = 17, "Shady side of pyramid" = 5), 
  init.angle = 315, 
  col = c("deepskyblue", "yellow", "yellow3"), 
  border = FALSE
)
```
Or you can use pandoc markdown for static figures with the following syntax:

`![optional caption that appears below the figure](figure url){alt='alt text for
accessibility purposes'}`

![You belong in The Carpentries!](https://raw.githubusercontent.com/carpentries/logo/master/Badge_Carpentries.svg){alt='Blue Carpentries hex person logo with no text.'}

## Math

One of our episodes contains $\LaTeX$ equations when describing how to create
dynamic reports with {knitr}, so we now use mathjax to describe this:

`$\alpha = \dfrac{1}{(1 - \beta)^2}$` becomes: $\alpha = \dfrac{1}{(1 - \beta)^2}$

Cool, right?

::::::::::::::::::::::::::::::::::::: keypoints

- Use `.md` files for episodes when you want static content
- Use `.Rmd` files for episodes when you need to generate output
- Run `sandpaper::check_lesson()` to identify any issues with your lesson
- Run `sandpaper::build_lesson()` to preview your lesson locally

>>>>

---
title: 'Multiple sequence alignment'
teaching: 0
exercises: 240
---

@@

In this short episode, we are exploring multiple sequence alignment (MSA). We are going to use [`mafft`](https://mafft.cbrc.jp/alignment/software/algorithms/algorithms.html) to align homologs of [RpoB](https://en.wikipedia.org/wiki/RpoB), the β subunit of the bacterial RNA polymerase. It is a long, multi-domain protein, suitable for showing issues related to MSA.

First, go to your home folder and create a phylogenetics folder

:::::::::::::::::::::::::::::::::::::::  challenge

## Exercise

Go to your home, use the `mkdir` command to make a `phylogenetics` subfolder, and move into it:

:::::::::::::::  solution

## Solution

```bash
$ cd ~
$ mkdir phylogenetics

```

:::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::::::::::::::::

Second, download a [fasta file](episodes/files/rpoB.fasta) containing 19 sequences of RpoB, gathered by aligning RpoB from *E. coli* to a very small database present at NCBI, the [landmark](https://blast.ncbi.nlm.nih.gov/smartblast/smartBlast.cgi?CMD=Web&PAGE_TYPE=BlastDocs) database. Move that file to your newly created `~/phylogenetics/` folder.

## Renaming sequences

Look at the accession ids of the fasta sequences: they are not very informative:

```bash
$ grep '>' rpoB.fasta | head -n 5
```

```output
>WP_000263098.1 MULTISPECIES: DNA-directed RNA polymerase subunit beta [Enterobacteriaceae]
>WP_011070610.1 DNA-directed RNA polymerase subunit beta [Shewanella oneidensis]
>WP_003114335.1 DNA-directed RNA polymerase subunit beta [Pseudomonas aeruginosa]
>WP_002228870.1 DNA-directed RNA polymerase subunit beta [Neisseria meningitidis]
>WP_003436174.1 MULTISPECIES: DNA-directed RNA polymerase subunit beta [Eubacteriales]
```

Keep the taxonomic name following the accession id, separated by `_`, using a bit of `sed` magic. Save the resulting file for later use, and show the headers again. 
```bash
$ sed "/^>/ s/>\([^ ]*\) \([^\[]*\)\[\(.*\)\]/>\\3_\\1/"  rpoB.fasta | sed "s/ /_/g" > rpoB_renamed.fasta
$ grep '>' rpoB_renamed.fasta | head -n 5
```

```output
>Enterobacteriaceae_WP_000263098.1
>Shewanella_oneidensis_WP_011070610.1
>Pseudomonas_aeruginosa_WP_003114335.1
>Neisseria_meningitidis_WP_002228870.1
>Eubacteriales_WP_003436174.1
```

It looks better.

## Alignment with progressive algorithm

Use `mafft` with a progressive algorithm to align the sequences.

:::::::::::::::::::::::::::::::::::::::  challenge

Use the FFT-NS-2 algorithm from mafft to align the renamed sequences. Also, record the time it takes for `mafft` to complete the task.

Hint: type `mafft` and try tab-complete to see all versions of `mafft`.

Hint: try the command `time`

:::::::::::::::  solution

```bash
$ time mafft-fftns rpoB_renamed.fasta > rpoB.fftns.aln
```

```output
...
Strategy:
 FFT-NS-2 (Fast but rough)
 Progressive method (guide trees were built 2 times.)

If unsure which option to use, try 'mafft --auto input > output'.
For more information, see 'mafft --help', 'mafft --man' and the mafft page.

The default gap scoring scheme has been changed in version 7.110 (2013 Oct).
It tends to insert more gaps into gap-rich regions than previous versions.
To disable this change, add the --leavegappyregion option.

mafft-fftns rpoB_renamed.fasta > rpoB.fftns.aln  1.02s user 0.52s system 104% cpu 1.471 total
```

The last line is the output of the `time` command. It took 1.47 seconds to complete this time.

:::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::::::::::::::::

## Alignment with iterative algorithm

Now use one of the supposedly better iterative algorithm of `mafft` to align the same sequences. Choose the [E-INS-i algorithm](https://mafft.cbrc.jp/alignment/software/algorithms/algorithms.html) which is suited for proteins that have highly conserved motifs interspersed with less conserved ones.

:::::::::::::::::::::::::::::::::::::::  challenge

Use the E-INS-i algorithm from mafft to align the renamed sequences. Also, record the time it takes for `mafft` to complete the task.

:::::::::::::::  solution

```bash
$ time mafft-einsi rpoB_renamed.fasta > rpoB.einsi.aln
```

```output
...
Strategy:
 E-INS-i (Suitable for sequences with long unalignable regions, very slow)
 Iterative refinement method (<16) with LOCAL pairwise alignment with generalized affine gap costs (Altschul 1998)

If unsure which option to use, try 'mafft --auto input > output'.
For more information, see 'mafft --help', 'mafft --man' and the mafft page.

The default gap scoring scheme has been changed in version 7.110 (2013 Oct).
It tends to insert more gaps into gap-rich regions than previous versions.
To disable this change, add the --leavegappyregion option.

Parameters for the E-INS-i option have been changed in version 7.243 (2015 Jun).
To switch to the old parameters, use --oldgenafpair, instead of --genafpair.

mafft-einsi rpoB_renamed.fasta > rpoB.einsi.aln  6.84s user 0.65s system 99% cpu 7.542 total
```

It now took 7.54 seconds to complete this time, i.e. about 5 times slower than with the progressive algorithm. It doesn't make a big difference now, but with hundreds of sequences it will make one.

:::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::::::::::::::::

## Alignment visualization

We want to compare the two alignments. Use `seaview`:

```bash
$ seaview rpoB.fftns.aln &
$ seaview rpoB.einsi.aln &
```

Arrange the two windows on top of each other. Change the fontsize (Props -> Fontsize) to 8 to see a larger portion of the alignment.

:::::::::::::::::::::::::::::::::::::::  challenge

Can you spot differences? Which alignment is longer?

Hint: try to scroll to position 800-900.

:::::::::::::::  solution

![](episodes/fig/seaview.png){alt='Alignments shown in seaview'}

::::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::::::::::::::::

If you have time over, spend it exploring the different options of Seaview.

### 2024-03-13 episodes/phylogenetics.Rmd
New episode of phylogenetics with NJ and ML.
Adaptation to Microbial Evolution and Epidemiology course; added pages, fused some

---
title: 'Phylogenetics'
teaching: 10
exercises: 20
---

@@

- How do you build a phylogenetic tree?

@@

- Learn the basic of phylogenetics tree building, taking the simplest of
the examples with a UPGMA tree.

@@

@@

- The simplest of the trees are distance-based.
- UPGMA works by clustering the two closest leaves and recalculating the 
distance matrix.

@@

>>>>

---
title: 'Phylogenetics'
teaching: 0
exercises: 240
---

@@

- How do you build a simple, distance-based phylogenetic tree?
- How do you build a phylogenetic tree with more advanced methods?
- How do you ascertain statistical support for phylogenetic trees?

@@

- Learn the basic of phylogenetics tree building, taking the simplest of the examples with a UPGMA tree.
- Learn how to use phylogenetics algorithms, neighbor-joining and maximum-likelihood.
- Learn how to perform and show bootstraps.

@@

## Introduction

In the previous episode, we inferred a simple phylogenetic tree using UPGMA, without correcting the distance matrix for multiple substitutions. UPGMA has many shortcomings, and gets worse as distance increases. Here we'll test Neighbor-Joining, which is also a distance-based, and a maximum-likelihood based approach with [IQ-TREE](http://www.iqtree.org/).

## Neighbor-joining

The principle of [Neighbor-joining method](https://en.wikipedia.org/wiki/Neighbor_joining) (NJ) is to start from a star-like tree, find the two branches that, if joined, minimize the total tree length, join then, and repeat with the joined branches and the rest of the branches.

To perform NJ on our sequences, we'll use the function inbuilt in Seaview. First (re-)open the alignment we obtained from `mafft` with the E-INS-i method.

```bash
seaview rpoB.einsi.aln &
```

In the Seaview window, select Trees -> Distance methods. Keep the BioNJ option ticked. BioNJ is an updated version of NJ, more accurate for longer distances. Tick the "Bootstrap" option and leave it at 100. We'll discuss these later. Then click on "Go".

What do you see? Is the tree rooted? Is the pattern species coherent with what you know of the tree of life? Any weird results?

Redo the BioNJ tree for the other alignment we inferred.

```bash
seaview rpoB.fftns.aln &
```

Do you see any differences between the two trees? What can you make out of it?

## Maximum likelihood

We will now use IQ-TREE to infer a maximum-likelihood tree of the RpoB dataset we aligned with `mafft` previously.

```bash
$ iqtree -s rpoB.einsi.aln -m TEST -B 1000
```

Here, we tell IQ-TREE to use the alignment `rpoB.einsi.aln`, and to test among the standard substitution models which one fits best (`-m TEST`). We also tell IQ-TREE to perform 1000 ultra-fast bootstraps (`-B 1000`). We'll discuss these later.

IQ-TREE is a very complete program that can do a large variety of phylogenetic analysis. To get a flavor of what it's capable of, look at its [extensive documentation](http://www.iqtree.org/doc/).

Have a look at the output files:

```bash
ls rpoB.einsi.aln.*
```

```output
rpoB.einsi.aln.bionj      rpoB.einsi.aln.iqtree     rpoB.einsi.aln.model.gz
rpoB.einsi.aln.ckp.gz     rpoB.einsi.aln.log        rpoB.einsi.aln.splits.nex
rpoB.einsi.aln.contree    rpoB.einsi.aln.mldist     rpoB.einsi.aln.treefile
```

There are two important files:  
* `*.iqtree` file provides a text summary of the analysis. 
* `*.treefile` is the resulting tree in [Newick format](https://en.wikipedia.org/wiki/Newick_format).

Now display the resulting tree in FigTree.

```bash
$ figtree rpoB.einsi.aln.treefile &
```

A window will pop-up and ask to provide a name for branch/node labels. Use e.g. 'bootstraps'.

The tree appears as unrooted. It is good practice to start by ordering the nodes and root it. Since we don't really know otherwise where the root is, we'll do a mid-point rooting to start with.

In the left menu, develop the 'Trees' menu. Tick the 'Root tree' and select 'Midpoint' from the drop-down menu. Tick the 'Order nodes' box and keep the 'increasing' option selected.

Scrutinize the tree. Is it different from the BioNJ tree generated in Seaview? How?

Redo a ML tree from the other alignment (FFT-NS) we inferred with `mafft` and display the resulting tree in FigTree

```bash
$ iqtree -s rpoB.fftns.aln -m TEST -B 1000
$ figtree rpoB.fftns.aln.treefile &
```

## Bootstraps

We have inferred four trees:
* Two based on the alignment generated from the E-INS-i algorithm, two from the FFT-NS.
* Two inferred with the BioNJ algorithm and two with the ML algorithm (IQ-Tree)

Along the way, we've generated bootstraps for all our trees. Now show them on all four trees.
* For the Seaview trees, tick the 'Br support' box
* For the trees shown in FigTree, develop the 'Branch labels' menu, tick the corresponding box and select 'bootstraps' from the 'Display' drop-down menu.

Compare all four trees. Do you find any significant differences?

Hint: what are *Glycine* and *Arabidopsis*? What about *Synechocystis* and *Microcystis*?

Hint: search for the accession number of the *Glycine* RpoB sequence on [NCBI](https://www.ncbi.nlm.nih.gov/search/). Any hint there?

The two plant RpoB sequences are from chloroplastic genomes, and thus branch together with the cyanobacterial sequences, in some of the phylogenies at least.

What about the support values for grouping these two groups? How high are they?

@@

- The simplest of the trees are distance-based.
- UPGMA works by clustering the two closest leaves and recalculating the distance matrix.
- Neighbor-joining is distance-based and fast, but not necessarily very accurate
- Maximum-likelihood is slower, but more accurate

@@

## Genetic drift

As a practical way to understand genetic drift, let's play with population size, selection coefficients, number of generations, etc.

- How does random genetic drift influence fixation of alleles?
- What is the influence of population size?

- Understand how population size influences the probability of fixation of an allele.
- Understand how slightly deleterious mutations may get fixed in small populations.

### Introduction

This exercise is to illustrate the concepts of selection, population size and
genetic drift, using simulations. We will use mostly 
[Red Lynx] by 
[Reed A. Cartwright](https://github.com/reedacartwright).

Another option is to
use a [web interface](https://phytools.shinyapps.io/drift-selection/) to the [R]
[learnPopGen](https://github.com/liamrevell/learnPopGen) package, but the last
one is mostly for biallelic genes (and thus not that relevant for bacteria).

Open now the [Red Lynx] website and get familiar with the different options.

You won't need the right panel (but feel free to explore). The dominance option
in the left panel won't be used either.

### Genetic Drift without selection

In the first part, you will only play with the number of generations, the
initial frequency and the population size.

- Lower the number of generations to 200. 
- Adjust the population size to 1000.
- Make sure the intial frequency is set to 50%.
- Run a few simulations (ca 20).

Did any allele got fixed? What is the range of frequencies after 200 generations?

In my simulations, no allele got fixed, the final allele frequencies range 
20-80%

Now increase the population to 100'000, clear the graph, and repeat the 
simulations. What's the range of final frequencies now?

In my simulations, no allele got fixed, and the final allele frequencies range
45-55%

Now decrease the population to 10 individuals, clear the graph and repeat
these simulations. What's the range of final frequencies now?

In my simulations, one allele got fixed quickly, the latest one was at
generation 100.

What do you conclude here?

It is clear that stochastic (random) variation in allele frequencies 
strongly affects the probability of fixation of alleles in small populations,
not so much in large ones.

### Genetic Drift with selection

So far we've only looked at how allele frequencies vary in the absence of 
selection, that is when the two alleles provide an equal fitness. What's the
influence of random genetic drift when alleles are not neutral?

The selection strength is equivalent to the selection coefficient, i.e.
how much higher relative fitness the new allele provides. A selection 
coefficient of 0.01 means that the organism harboring the new allele has a 
1% increased fitness.

- Increase the number of generations to 1000.
- Set the selection strength to 0.01 (1%).
- Set the population size first to 100'000, then to 1000, then to 100.

How long does it take for the allele to get fixed - in average - with
the three population sizes?

About the same time, but the trajectories are much smoother with larger
populations. In the small population, it happens that the beneficial allele
disappears from the population (although not often).

### Fixation of slightly deleterious alleles in very small populations

We're now simulating what would happen in a very small population (or a 
population that undergoes very narrow bottlenecks), when a gene mutates. We'll
have a very small population (10 individuals), a selection value of -0.01 
(the mutated allele provides a 1% lower fitness), and a 10% initial frequency, 
which corresponds to one individual getting a mutation:

- Set the population to 10 individuals.
- Set generations to 200.
- Set initial frequency to 10%.
- Set the selection strength to -0.01.

Run many simulations. What happens?

Most new alleles go extinct quickly. In my simulations, I usually get one
of the slightly deleterious mutations fixed after 20 simulations.

:::::::::::::::::::::::::::::::::
::::::::::::::::::::::::::::::::::::::::::::::::

That's it for that part, but feel free to continue playing with the different 
settings later.

- Random genetic drift has a large influence on the probability of fixation of
alleles in small populations, even for non-neutral alleles.
- Random genetic drift has very little influence of the probability of fixation of
alleles in large populations.
- Slightly deleterious mutations can get fixed into the population through
random genetic drift, if the population is small enough and the selective value
is not too large.

<!-- References -->

[Red Lynx]: https://cartwrig.ht/apps/redlynx/

### 2024-03-13 episodes/sequence-assembly.Rmd
Adaptation to Microbial Evolution and Epidemiology course; added pages, fused some

---
title: 'Sequence assembly'
teaching: 10
exercises: 30
---

@@

To run SKESA we will use the skesa command with a number of option that we will explore later on. We can start the loop with the assemblies with:

```bash
$ for sample in ERR029206 ERR029207
do
skesa \
--cores 2 \
--memory 8 \
--fastq "${sample}"_1.fastq.gz_trim.fastq "${sample}"_2.fastq.gz_trim.fastq 1> ../../results/assembly/"${sample}".fasta
done
```

@@

@@

>>>>

---
title: 'Sequence assembly'
teaching: 0
exercises: 240
---

@@

To run SKESA we will use the skesa command with a number of option that we will explore later on. We can start the loop (again, a short loop) with the assemblies with:

```bash
$ for sample in ERR029206
do
skesa \
--cores 2 \
--memory 8 \
--fastq "${sample}"_1.fastq.gz_trim.fastq "${sample}"_2.fastq.gz_trim.fastq 1> ../../results/assembly/"${sample}".fasta
done
```

@@

:::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::: instructor

@@

:::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::

### 2024-03-13 episodes/snps-phylogeny.Rmd
Adaptation to Microbial Evolution and Epidemiology course; added pages, fused some

---
title: 'SNP phylogeny'
teaching: 15
exercises: 60
---

@@

@@

Snippy is fast but a single run still takes about 2-3 minutes. We will therefore tell snippy to run all the samples after each other. However this time we can not use a wildcard to do so. We will instead run all the samples in a loop.

```bash
$ cd ../data/trimmed_fastq/
$ for sample in ERR026473 ERR026474 ERR026478 ERR026481 ERR026482 ERR029206 ERR029207
do
snippy  --outdir ../../results/snps/"${sample}" --ref ../GCF_000195955.2_ASM19595v2_genomic.fna --R1 "${sample}"_1.fastq.gz_trim.fastq --R2 "${sample}"_2.fastq.gz_trim.fastq
done
```

```bash
$ head -n10 ~/molepi/results/snps/ERR029207/snps.tab 
```

```output
CHROM	POS	TYPE	REF	ALT	EVIDENCE	FTYPE	STRAND	NT_POS	AA_POS	EFFECT	LOCUS_TAG	GENE	PRODUCT
NC_000962.3	1849	snp	C	A	A:215 C:0
NC_000962.3	1977	snp	A	G	G:113 A:0
NC_000962.3	4013	snp	T	C	C:128 T:0
NC_000962.3	7362	snp	G	C	C:131 G:0
NC_000962.3	7585	snp	G	C	C:129 G:0
NC_000962.3	9304	snp	G	A	A:123 G:0
NC_000962.3	11820	snp	C	G	G:157 C:0
NC_000962.3	11879	snp	A	G	G:90 A:0
NC_000962.3	14785	snp	T	C	C:174 T:0
```

This list gives us information on every SNP that was found by snippy when compared to the reference genome. The first SNP is found at the position 1849 of the reference genome, and is a C in the H37Rv (reference strain) and an A in isolate ERR029207. There is a high confidence associated with this SNP: an A has been found 225 times in the sequencing reads and never a C at this position.

@@

Find out how many SNPs were identified in the *M. tuberculosis* isolates when compared to H37Rv. 
Hint: The .txt file in the snippy output contains summary information

@@

@@

Phylogenetic trees have been discussed during the lectures. We will here establish a phylogenetic tree from the file 'core.aln' with [IQ-TREE](https://www.iqtree.org/). IQ-TREE is a phylogeny software based on the maximum-likelihood principle. IQ-TREE has been widely used and cited. There is a very wide range of parameters that need to be chosen, such as nucleotide or amino-acid substitution models.

>>>>

---
title: 'Mapping'
teaching: 0
exercises: 240
---

@@

::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::: instructor

@@

We need to activate a separate environment to run snippy:

```bash
$ mamba activate snippy
```

Snippy is fast but a single run still takes about 2-3 minutes. We would therefore tell snippy to run all the samples after each other. However this time we cannot use a wildcard to do so. We would instead run all the samples in a loop. The "real" loop is commented out, and we use one that runs only the sample we've analyzed so far.

```bash
$ cd ../data/trimmed_fastq/
$ #for sample in ERR026473 ERR026474 ERR026478 ERR026481 ERR026482 ERR029206 ERR029207
$ for sample in ERR029206
do
snippy --ram 1 --outdir ../../results/snps/"${sample}" --ref ../GCF_000195955.2_ASM19595v2_genomic.fna --R1 "${sample}"_1.fastq.gz_trim.fastq --R2 "${sample}"_2.fastq.gz_trim.fastq
done
```

Here, we provide snippy with an output folder (`--outdir`), the location of the reference genome (`--ref`), and the trimmed read files for each end of the pair (`--R1` and `--R2`). We also indicate that snippy should use 1 Gb of memory (`--ram 1`, very specific issue to this VM).

```bash
$ head -n10 ~/molepi/results/snps/ERR029206/snps.tab 
```

```output
CHROM	POS	TYPE	REF	ALT	EVIDENCE	FTYPE	STRAND	NT_POS	AA_POS	EFFECT	LOCUS_TAG	GENE	PRODUCT
NC_000962.3	1849	snp	C	A	A:217 C:0
NC_000962.3	1977	snp	A	G	G:118 A:0
NC_000962.3	4013	snp	T	C	C:176 T:0
NC_000962.3	7362	snp	G	C	C:156 G:0
NC_000962.3	7585	snp	G	C	C:143 G:0
NC_000962.3	9304	snp	G	A	A:127 G:0
NC_000962.3	11820	snp	C	G	G:164 C:0
NC_000962.3	11879	snp	A	G	G:125 A:0
NC_000962.3	14785	snp	T	C	C:163 T:1
```

This list gives us information on every SNP that was found by snippy when compared to the reference genome. The first SNP is found at the position 1849 of the reference genome, and is a C in the H37Rv (reference strain) and an A in isolate ERR029206. There is a high confidence associated with this SNP: an A has been found 217 times in the sequencing reads and never a C at this position.

Now download a [reduced version of the result of snippy](files/snps.tar.gz) on the other samples, save it in the `~/molepi/results/` folder. We will also rename the existing `snps` folder to save it.

```bash
cd ~/molepi/results/
mv snps snps_partial
mkdir snps
cd snps
tar xvzf ../snps.tar.gz
ls
```

```output
ERR026473     ERR026482
ERR026474     ERR029206
ERR026478     ERR029207
ERR026481
```

The folders don't contain everything that is output by `snippy`, to save space. But it is enough to run `snippy-core`.

```bash
ls ERR026473
```

```output
ref.fa          snps.filt.vcf   snps.tab
ref.fa.fai      snps.gff        snps.txt
snps.aligned.fa snps.html       snps.vcf
snps.bam.bai    snps.log        snps.vcf.gz
snps.bed        snps.raw.vcf    snps.vcf.gz.csi
snps.csv        snps.subs.vcf
```

@@

Find out how many SNPs were identified in the *M. tuberculosis* isolates when compared to H37Rv. 
Hint: The `.txt` file in the snippy output contains summary information:

@@

Let's deactivate the specific snippy `env` before we move on:

```bash
$ mamba deactivate snippy
```

@@

Phylogenetic trees have been discussed during the lectures. We will here establish a phylogenetic tree from the file 'core.aln' with [IQ-TREE](http://www.iqtree.org/). IQ-TREE is a phylogeny software based on the maximum-likelihood principle. IQ-TREE has been widely used and cited. There is a very wide range of parameters that need to be chosen, such as nucleotide or amino-acid substitution models.
